Staged Abstract Interpreters

### 2019-08-15 paper/20_prelim.tex
misc
update

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.

\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  for { x <- xs; y <- ys } yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

The Scala compiler will translate the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}. The last binding
in the @for@-comprehension is translated into a @map@, where the expression of
@yield@ becomes the body expression of that @map@ application. The foregoing
bindings in the comprehension are all translated into calls of @flatMap@.

\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object @List[_]@ encapsulates the data
internally.  Therefore it only exposes the simplified version of
@flatMap@, where the monad @M[A]@ is not introduced as a function
argument. The trait @MonadOps@ (Figure \ref{fig:monadops}) defines the
simplified version of monadic operations that are necessary for
@for@-comprehension. The conversion between @Monad@ and @MonadOps@
can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax
and a few of monads and monad transformers such as @ReaderT@,
@StateT@, and @ListT@ to write our interpreters. The implementation
of monads and monad transformers essentially borrows the ground-truth
from Haskell.

@@

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program runtime.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.

\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
First, two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.

\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic is handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.

\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.

%\vspace{-1em}

\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
type @Env@ and @Store@ is derived automatically. To assure the freshness of
address allocation, we use @Int@ and always return successor of the size of
current store. A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final grounded
result of the interpreter is a pair of values and stores, where the @Value@ and
@Store@ are next-stage objects, as annotated by @R@. We also define a standard
fixed-point combinator to close the open-recursive function @ev@.

\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

\paragraph{Unstaged Monads}
For the purpose of concrete interpretation, the monad needs to model read
effect and state effect, which correspond to environment and store,
respectively. Following the monad transformer approach
\cite{DBLP:conf/popl/LiangHJ95}, we use a @ReaderT@ and @StateT@ monad
transformer to compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, thus
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by type
@Env@, @StateT@ is parameterized by type @Store@, and the inner-most monad @IdM@
is simply an identity monad.

\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produecs a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.

\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations operates on the value domain can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. Because lambda terms are data but also part of the
control flow, although the way we treat them is simple now, but will be very
different when staging and abstract interpretation is involved. The method
@close@ takes a lambda term and an environment, and produces the
defunctionalized representation of closures.

\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

To apply a function, @ap_clo@ takes a function value @rator@ and an argument
@rand@, extracts the lambda term and environment enclosed in @rator@, allocates
the addresses and updates the environment and store accordingly. Finally, it
evaluates the body expression of the lambda term under the new environment and
store.

\begin{lstlisting}
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

The top-level @run@ method invokes the fixed-pointer operator directly,
given the expression @e@, initial environment $\rho_0$ and store $\sigma_0$.

\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ$_0$)(σ$_0$)
\end{lstlisting}

>>>>

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  for { x <- xs; y <- ys } yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

The Scala compiler will translate the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}. The last binding
in the @for@-comprehension is translated into a @map@, where the expression of
@yield@ becomes the body expression of that @map@ application. The foregoing
bindings in the comprehension are all translated into calls of @flatMap@.
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object @List[_]@ encapsulates the data
internally.  Therefore it only exposes the simplified version of
@flatMap@, where the monad @M[A]@ is not introduced as a function
argument. The trait @MonadOps@ (Figure \ref{fig:monadops}) defines the
simplified version of monadic operations that are necessary for
@for@-comprehension. The conversion between @Monad@ and @MonadOps@
can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax
and a few of monads and monad transformers such as @ReaderT@,
@StateT@, and @ListT/SetT@ to write our interpreters. The implementation
of monads and monad transformers essentially borrows the ground-truth
from Haskell.

@@

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program runtime.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
First, two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic is handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
type @Env@ and @Store@ is derived automatically. To assure the freshness of
address allocation, we use @Int@ and always return successor of the size of
current store. A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final grounded
result of the interpreter is a pair of values and stores, where the @Value@ and
@Store@ are next-stage objects, as annotated by @R@. We also define a standard
fixed-point combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

\paragraph{Unstaged Monads}
For the purpose of concrete interpretation, the monad needs to model read
effect and state effect, which correspond to environment and store,
respectively. Following the monad transformer approach
\cite{DBLP:conf/popl/LiangHJ95}, we use a @ReaderT@ and @StateT@ monad
transformer to compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, thus
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by type
@Env@, @StateT@ is parameterized by type @Store@, and the inner-most monad @IdM@
is simply an identity monad.
\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produecs a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations over the value domain can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. Because lambda terms are data but also part of the
control flow, although the way we treat them is simple now, but will be very
different when staging and abstract interpretation is involved. The method
@close@ takes a lambda term and an environment, and produces the
defunctionalized representation of closures.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

To apply a function, @ap_clo@ takes a function value @rator@ and an argument
@rand@, extracts the lambda term and environment enclosed in @rator@, allocates
the addresses and updates the environment and store accordingly. Finally, it
evaluates the body expression of the lambda term under the new environment and
store.
\begin{lstlisting}
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

The top-level @run@ method invokes the fixed-pointer operator directly,
given the expression @e@, initial environment $\rho_0$ and store $\sigma_0$.
\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ$_0$)(σ$_0$)
\end{lstlisting}

### 2019-08-15 paper/30_stageconc.tex
update

A classic example for introducing multi-stage programming is the power function
that computes $b^x$, which is usually implemented as a recursive function:

\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

If @x@ is a value know at current stage, we may specialize the power function to
some value @x@ -- by unrolling the recursive calls. In LMS, this is fulfilled by
first adding the @Rep@ type to the variable known at next stage. In this case,
@b : Rep[Int]@ is know later, and @x : Int@ is know currently, as shown in the
below code recipe.
The way we use this @power@ function is to create a @DslDriver@ and override the
@snippet@ method give the currently know value @x@ (5 in this example).

\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5) // specialize the power to b^5
  }
\end{lstlisting}

@@

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. But now the monads operate on staged values, for
brivity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that objects of a case class @ReaderT@/@StateT@ are not staged, the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged. The following code snippet shows
the idea (we use light gray to highlight what have been changed to staged
version except naming):

\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a current-stage-known value, so
that all the invocation of @flatMap@s can be completed before generating code.
The fact that we only staged the data but not the monadic computation or monadic
values, is the reason that we can peel of the monad stack in the generated code.
Now the @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions.

\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

TODOTODO====

We use @RepStateT@ as an example to give a sketch of how the @Rep@ versions of monad
transformers look like. The generic type @M[_]@ is a @RepMonad@ and used as the
inner monad, which means the pair @(A,S)@ is also staged when appears inside of the monad
@M[_]@. The @flatMap@ is similar to the unstaged one, but the function @f@ takes
a @Rep[A]@ value. Again, @f@ is not a next-stage function, but a current-stage
function that transforms a next-stage value of @Rep[A]@ to @StateT[M,S,B]@.
%\todo{highlight the changes in the code}

Readers may notice that the conversion between unstaged monads and staged monads
is simply changing the unstaged data to staged data, which in fact can be
achieved without modifying the implementation of monads. This is true so far, but
as we will see, it is not that easy when introducing the nondeterminism monad
(@ListT@) for abstract interpreters, as the whole list will be a next-stage
value instead of merely the elements in the list. For this reason, we explicitly
distinguish and present the two versions of monads, as a more consistent
approach.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to let them works on staged data.
As we mentioned before, we elaborate on how to handle functions and applications in
detail. One of the desired goals is to eliminate the monadic layers through
staging, this becomes a little bit tricky for closures because we actually need 
 to compile the lambda term instead of generating a @CloV@ value, but the @ev(e)@
returns a monad and prevents the specialization. To make this happen, when
closing a lambda term, we need to specialize the body expression by
\textit{collpasing} the @AnsM@ monads to normal values. The
following function @close@ illustrates the idea: @f@ is a current stage function
but takes two next-stage values @v@ and $\sigma$, inside of which, we eagerly
collapse the monads to values by @ev(e)(ρ_*)(σ_*)@, which has type
@Rep[(Value, Store)]@. Then we generate the next-stage value of the compiled
closure using @emit_compiled_clo@, containing the compiled closure.

@@

>>>>

A classic example for introducing multi-stage programming is the power function
that computes $b^x$, which is usually implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

If @x@ is a value know at current stage, we may specialize the power function to
some value @x@ -- by unrolling the recursive calls. In LMS, this is fulfilled by
first adding the @Rep@ type to the variable known at next stage. In this case,
@b : Rep[Int]@ is know later, and @x : Int@ is know currently, as shown in the
below code recipe.
The way we use this @power@ function is to create a @DslDriver@ and override the
@snippet@ method give the currently know value @x@ (5 in this example).
\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5) // specialize the power to b^5
  }
\end{lstlisting}

@@

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. But now the monads operate on staged values, for
brivity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that objects of a case class @ReaderT@/@StateT@ are not staged, the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged, i.e., @R@ and @A@ in the
reader monad, @S@ and @A@ in the state monad. The following code
snippet shows the idea. We use \hl{light gray} to highlight where @Rep@ type is added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a current-stage-known value, so
that all the invocation of @flatMap@s can be completed before generating code.
The fact that we only stage data but not monadic computation or monadic
values, is the reason that we can peel of the monad stack in the generated code.
Now the @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can be
achieved without modifying the implementation of monads. This is true so far, but
as we will see, it is not straightfoward when for nondeterminism monad
(@SetT@) in abstract interpreters. Because not only the elements in the set are staged,
but the whole set is also staged, and we have no knowledge about how many elements
in the set. In this section, we explicitly distinguish the two versions of monadic interfaces,
later we will instantiated staged set monad using a different strategy.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.
As we mentioned before, we will take more care on how functions and
applications are handled. The problem here is what should we do for
lambda terms when staging? We cannot create a next-stage
defunctionalized closure for it, because it means we still need to
access the interpreter at the next stage. The desired goal is to eliminate the
interpretation overhead staging, so the right thing is compiling the lambda term
and its environment to a next-stage Scala function. To do this, we need to
recursively call the interpreter function, i.e., the @ev@ argument passed to @close@.
The following code implements the idea:

@@

We have created a current function @f@ that takes a next-stage value
and store, and returns staged value. Then we use @emit_complied_clo@
to delay @f@ to a next Scala function, represented by the type
@Rep[Value]@. Inside of @f@, we

To make this happen, when
closing a lambda term, we need to specialize the body expression by
\textit{collpasing} the @AnsM@ monads to normal values. The
following function @close@ illustrates the idea: @f@ is a current stage function
but takes two next-stage values @v@ and $\sigma$, inside of which, we eagerly
collapse the monads to values by @ev(e)(ρ_*)(σ_*)@, which has type
@Rep[(Value, Store)]@. Then we generate the next-stage value of the compiled
closure using @emit_compiled_clo@, containing the compiled closure.
